{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-09-24 Docs/manuscript.tex
more edits, fixed plots

%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{July 2023}
\fancyhead[RO]{Manuscript}
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

% Just jotting down and putting some bits you have for abstract for now. 
\begin{abstract}
Climate change forecasting .... plasticity.... adaptation. Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because... 
\end{abstract} % seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019

\section{Introduction}
%emw: FYI this is a first draft -- meaning it still has lots of words, sentences and other bits to fix; but best to do that as we go. 
%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much? I suspect it may be contested. I also wonder if my advice about definitions was wrong ... but let's definitely save that text for now. 
%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.

% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. 
Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}. In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation \citep{chevin102, snell18}.

@@

Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies  (i.e., studies---often for forestry purposes---where researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype) \citep{AitkenBemmels16, Alberto13}. This work has established common clines in fall phenology, as source locations with shorter growing season (poleward and higher elevations) exhibit earlier stopping of growth (e.g., budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold that trigger budset), driven by local adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears far more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows high enough levels of adaptation to matter biologically.  % alinaJuly20: im not sure what important levels of adaptation means here, %emwaug1: I meant that the 'amount' of local adaptation is high enough to matter (contrasting with the reality that if you have a big enough sample you can often find a small difference that is 'significant' but it might not matter biologically).

Results, however, have not always been consistent. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have not found any trends in budburst along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to some debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they could vary by continent.

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional provenance trials, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common garden differences in study design---especially studied species---may complicate understanding what underlies potential trends across continents, with European studies focusing more on deciduous angiosperms and North American studies focusing often on evergreen gymnosperms.

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies.} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}and ClimateWNA \citep{wang2016}.

To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C).

@@

\section{Results}
Our final dataset included seven angiosperm and eight gymnosperm species from 17 studies, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species also had fall event information available.

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT:-0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) advanced strongly with provenance latitude (3.16 days/degree [2.87-3.45]) and the decrease in MAT (4.78 days/$^{\circ}$C [4.1 - 5.4]), meaning fall events were earlier where provenance MAT was lower (Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts local adaptation. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe. In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}. While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests overall no general trend for clines in spring phenology. While not a direct test, our results suggest that spring events are highly plastic; they thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures, in line with most current climate change impact studies \citep{IPCC22}

Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend (albeit a much weaker one than for fall phenology). Clines with MAT diverged between angiosperms and gymnosperms. Combined with our finding of stronger clines for fall phenology in North America, our results suggest that data on more species tested across continents could yield important insights into plasticity versus adaptation limits for spring and fall phenology, which would greatly aid forecasting.

We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in months close to the events, at least in the narrow range of northern temperature and boreal forests we studied.

\subsection{Variation across continents and species groupings}

Our results indicated that the effects of provenance latitude and MAT on fall events were similar across angiosperm and gymnosperm species: Fall events delay as MAT increases for both leaf types, slightly more so for gymnosperms, explained by sensitivity to day-length signals. However, the two leaf types exhibited slightly diverging trends in spring events:  advancing in angiosperms and delaying in most gymnosperms as MAT increases. Such differences make sense as timing of budburst and leafout varies by leaf type. Gymnosperms photosynthesize without leafout and thus generally leafout later than angiosperms, after much risk of spring frost \citep{panchen14}. Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

While we noticed more prominent clines in fall events in North America, there was no strong evidence of overall continental differences in spring events, which might again suggest a higher degree of plasticity in spring events. Meanwhile, most North American gardens and provenances included in this study were limited to the Pacific Northwest region, which might mask the kind of climate pattern we were hoping to see from the entirety of North America (Fig. \ref{figure:map_gardens}).

\subsection{Adaptation in spring vs fall phenology: Implications for climate change responses}

Our results suggest that current advances in spring phenology may continue, but delays in fall phenology could depend on the speed of adaptation. While some previous chamber studies have found potential adaptation, these could have been influenced by chilling at the provenance site \citep{ettinger20}. However, field studies and common garden experiences do not have this effect. Spring events from the studies we looked at appeared to exhibit high plasticity, while fall events showed strong clines, which is consistent with previous work \citep{AitkenBemmels16, Alberto13}. These findings have important implications for growing discussion of how shifting phenology and tree growth relate as our results suggest there are strong limits on shifts in fall phenology \citep{zohner23, zani20}.

It is worth noting that these trends likely vary by species and depend on frost tolerance strategy. If current advances in spring phenology continues, it could suggest shifts in spring frost. In temperate and boreal environments, the optimal timing for spring growth initiation is determined by a balance of avoiding late spring frost and having a long growing season \citep{alberto11, lenz16, allevato19}. Utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15} and the ones with suboptimal shade tolerance \citep{richardson09}. Further, plants that budburst earlier can take advantage of a longer growing season and avoid competition \citep{guo22}. However, differences across species means that this will not impact all species the same. While fall event always exhibited co-gradient relationships (i.e. events advanced as we moved northward or up in elevation), spring events were more varied among species. The implication of such variations could be that spring leafout sets up flowering, and thus, some counter-gradient could better aid gene glow over larger geographical regions.

All of above are critical to predicting not just phenological change, but range expansions with climate change: An accurate understanding of how environmental conditions might influence species distribution at a continental scale is critical for future range shift prediction. Implications of small differences in spring phenology could mean different levels of gene flow.

Our observations reveal that although temperature by itself could serve as a reasonable indicator for predicting when fall events occur across various locations, it's crucial to acknowledge that the degree of adaptation dictated by climate change differs across different geographical areas \citep{Loarie09}. High-latitude sites, for example, are likely restricted by the impact of photoperiod \citep{Gill15}. Diverse mechanisms that govern the timing of leaf senescence at the ecosystem level hold the potential to exert significant influence over not only the interactions occurring within plant communities but also the intricate process of carbon storage within ecosystems. With the anticipated escalation of global temperatures in the forthcoming century, these mechanisms are poised to play a pivotal role in shaping the dynamics of plant communities and the overall carbon balance of ecosystems.

>>>>

%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2023}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\linenumbers
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
\doublespacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\begin{abstract}
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through both inter-annual flexible variation versus through adaptation to local conditions. Yet we often lack this information entirely, or find conflicting evidence across studies. This is the case for one of the most reported and consistent impacts of anthropogenic climate change---earlier spring phenology, which is also one of the most critical to forecasting given its role in carbon sequestration. Evidence from some common garden studies suggest important provenance effects, which suggest local adaptation in the underlying cues of spring phenology and supports findings for fall events, while other studies find no evidence. Here we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 50 years of common garden studies on tree phenology we may still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\section{Introduction}

%emw: I noticed a lot of papers on local adaptation don't always use the term 'local adaptation' so we may want to avoid using it too much? I suspect it may be contested. I also wonder if my advice about definitions was wrong ... but let's definitely save that text for now. 
%emw: We need to 'funnel' in our intros -- until we submit to a plant journal focusing on plants too early is getting too narrow too fast.

% To optimize ecosystem services for habitat conservation and understand how plants cope with environmental variations, predicting the biological impacts of climate change is of critical importance \citep{botero15}. One of the most important parts of accurate prediction is understanding the relative importance of local adaptation versus plasticity \citep{chevin10}. 
Predicting the biological impacts of climate change has made understanding how organisms cope with environmental variation more urgent \citep{botero15}. In particular, the relative importance of plasticity versus genetic adaptation is vital for prediction \citep{chevin10}, with plasticity expected to allow species to shift more rapidly with climate change than environmental responses based on local adaptation, but possibly limiting responses after the limits of plasticity are reached \citep{chevin102,snell18}.

@@

Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenology and phenotype \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and is determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

These contrasting studies highlight how inconsistent evidence for adaptation in tree spring phenology has been. Studies have documented provenance differences of 2-4 days per degree latitude in spring phenology for some species (\emph{Picea abies} in \citealp{sog08} and \emph{Quercus petraea} in \citealp{deans96}) while others have failed to find similar trends along latitudinal gradients (for example, \emph{Picea sitchensis} in \citealp{mimura07}, \emph{Picea glauca} in \citealp{Li97}, and \emph{Populus balsamifera} in \citealp{farmer93}). This has led to debate over the prevalence and importance of adaptation in spring tree phenology. While clines of spring phenology have been found in studies in both Europe \citep{sog08,deans96,von95} and North America \citep{rossi15, soo13, hannerz99}, they may be more studies and debated in Europe \citep{deans96,vitasse2009,Basler:2012}, raising the possibility that they could vary by continent. %EMWSep4: Check deans96 is accurate ref I added. %alinaSep22: Hi lizzie I am not sure how you are trying to phrase "they may be more studies and debated in Europe" but I think it's slightly off. could you fix it?

Continental differences in patterns of adaptation versus plasticity could be driven by climatic differences, especially as North American springs are more variable across years than European ones \citep{tward21,zohner2017,schwartz00}. Such high temporal variability means that distant sites can effectively experience the same spring climate, but in different years. Studies of spring phenology in arboreta suggest cues for budburst may vary depending on continental climate \citep{zohner2017}, but are poorly controlled compared to traditional common garden studies, making them difficult to use for tests of plasticity versus adaptation \citep{gauzere2020}. Even for more carefully designed common gardens, differences in species studied or other differences in design may complicate understanding what underlies potential trends across continents.  %EMWSep4: Moved last point to discussion.

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig. \ref{figure:map_gardens} \& Supplement Table.\ref{supplement-table:all_studies}). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}. % emwJun4 -- you need to use ` to open single quotes in Latex. 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.\ref{supplement-table:all_studies} in Supplement for information on selected studies. Note: map lines delineate study regions and do not necessarily depict accepted national boundaries} 
    \label{figure:map_gardens}
\end{figure}
\newline
To understand how climatic differences, in addition to geographical differences, shape local adaptation in spring events we extracted several types of climate data using information about provenance latitude, longitude, and elevation from original publications. We estimated the mean annual temperature (MAT) from 1960-1991 for each provenance using the Climate Information Tool by Food and Agriculture Organization of the United Nations \citep{FAO2022}. We verified our calculated MAT was was similar to MAT calculated using ClimateWNA \citep{wang2016}, a source used in previous analyses.

To examine more explicitly climate near spring events, we used gridded daily temperature data for 2011 to 2020 for all European and North American provenances and gardens from E-OBS and the `daymetr' package in R respectively \citep{cornes2018,hufkens2018}. Using these data we estimated how much the daily temperatures overlapped between garden and provenance locations, which we call `climate overlap.' For this we used the `overlap' package in R to calculate the percentage overlap of the daily temperature of each provenance and their corresponding gardens in spring months (March to May) from 2011 to 2020. % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C). % See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.

@@

\section{Results}
Our final dataset included seven deciduous angiosperm and eight evergreen gymnosperm species from 17 studies and 19 gardens, encompassing 384 North American provenances and 101 European provenances, with observations from 1962 to 2019. Seven species (five in North America and two in Europe) also had fall event information available. Most species in North American gardens were gymosperms (7/11 species) while most species in European gardens were angiosperms (3/4 species).

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: 0.10 days/degree [-0.05 - 0.25]; MAT: -0.11 days/$^{\circ}$C [-0.34 - 0.12]) (Fig. \ref{figure:springfall_latmat}, Table. \ref{supplement-table:model_spring_lat} \& \ref{supplement-table:model_spring_mat} in Supplements), nor within North America (latitude: 0.10 days/degree [-0.06 - 0.26]; MAT: -0.09 days/$^{\circ}$C [-0.36 - 0.18]) or Europe (latitude: 0.10 days/degree [-0.23 - 0.42]; MAT: -0.16 days/$^{\circ}$C [-0.55 - 0.23]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

In contrast, fall events (e.g., budset, leaf senescence, leaf abscission) were earlier at more northern, cooler MAT sites (that is, they advanced strongly with provenance latitude: 3.16 days/degree [2.87-3.45], and with decreasing MAT:4.78 days/$^{\circ}$C [4.1 - 5.4], Fig. \ref{figure:springfall_latmat}, Table.\ref{supplement-table:model_fall_lat} \& \ref{supplement-table:model_fall_mat} in Supplements ). This relationship, however, was observed mostly in North America where fall events advanced 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT ($^{\circ}$C), whereas in Europe these relationships were weaker: advance of 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT ($^{\circ}$C) (Fig. \ref{figure:continental_spptype_effect}A).

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig. \ref{supplement-figure:overlap_scatterplot} in Supplements). We observed very weak effects of climate overlap on spring events (0.01 [0.02 - 0.03] days per one percent increase in climate overlap), nearly identical across angiosperms (0.02 [0.00 - 0.05]) and gymnosperms (0.04 [0.00 - 0.09]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing 0.72 [0.51 - 0.92] days per one percent decline in climate overlap) (Fig.\ref{supplement-figure:overlap_posterior} in Supplements).

@@

Overall, our results demonstrated inconsistent and weak clines in spring events across North America and Europe. In contrast, fall events generally showed much stronger clines, especially in North America, and in support of many previous studies \citep{AitkenBemmels16, Alberto13}. While previous studies have suggested spring events are far more plastic compared to fall events \citep{Li97,farmer93,mimura07}, our study provides the first major test of this across continents and species and suggests no general trend for clines in spring phenology. Our results thus predict that warming springs will continue to be tracked more closely phenologically by trees than warming fall temperatures \citep{IPCC22}

Trends between spring phenology and latitude or MAT were weak, but suggested the possibility of diverging results that could mute an overall trend---albeit a much weaker one than for fall phenology. We found angiosperm versus gymnosperm species diverged in their clines with MAT. Combined with our finding of much stronger clines for fall phenology in North America, these results support the idea of potential variation across continents and/or species type that may underlie the debate in whether spring events show important clinal variation. As clines with spring events were very weak, however, and gardens almost always focus on only one species, understanding these diverging results well enough to aid forecasting would take significant additional investment in common garden studies. %EMWSep4 -- we need to be careful in how many times we say 'we need more data' -- it's a common need, so it's better to focus on specifics of what the data suggests, or doesn't and say we need more data -- at most once in a paper. More on this below.

We found the coarse metrics of provenance latitude and MAT were generally good predictors of phenology, performing better than our more complex and data-rich metric of climate overlap. Latitude and MAT appeared to well represent how similar the climates are between the provenances and gardens in the temperate and boreal forest species we studied, yielding similar results metrics calculated specifically in the spring with daily climate data. %EMWSep4 -- also avoid repeating how narrow the data is etc. -- it's good to point out limitations, but repeating them too much is not helpful -- this is the biggest dataset for this question on earth, we need to focus on what we learned from it, not so much on the limitations. Also, Europeans would not call all of the forests northern.

\subsection{Variation across continents and species types}

Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by their varying investment strategies. As gymnosperms photosynthesize without leafout they generally leafout much later than angiosperms, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk more than angiosperms. In contrast, angiosperms may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require More information on frost risk and forest community assembly from across the provenance locations. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

In our dataset, more data on fall events were reported in North America. In Europe, only two gardens reported fall event metrics, one studying \emph{Fagus sylvatica} and one studying \emph{Fraxinus excelsior}. Fall events were earlier in colder, more northern latitudes in both continents, which is consistent with previous literature \citep{AitkenBemmels16, Alberto13}. North American studies of both gymnosperm and angiosperm species reported strong clines, and most often focused on budset, while the two European gardens reported data on `leaf senescence' (Supplement Table.\ref{supplement-table:all_studies}). Leaf senescence was visually inspected based on the percentage of the tree crown that had changed color when the assessor stood on the south-facing side of each tree \citep{rosique22}. We note that budset is most closely related to growth cessation and tied to photoperiod signals, making it generally considered a more robust indicator of when plants stop investing in growth and more genetically controlled than leaf senescence \citep{Alberto13}.

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}. Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig. \ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g.,\emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly.

\subsection{Adaptation in tree phenology: Implications for climate change responses}

%EMWSep4 -- good stuff here! I tried to refocus on spring events some as our data is focused there. And I limited the amount of hand-wavey stuff we discussed as reviewers only usually want so much extrapolation of your results (so I cut some of the ending stuff). 
Our results suggest that current advances in spring phenology will likely continue, but predicting any shifts due to provenance effects will require new data across more species. Plastic responses to warming, as our results suggest for spring events, mean species can adjust to the shifting environments of climate change---up to some point. Beyond that point theory suggests plastic traits may limit adaptation to climate change \citep{chevin10}. This may mean species will be replaced when their plastic spring phenological responses are no longer adequate, or shifting reaction norms could lead to adaptation \citep{gauzere2020}.

Predicting this will require better understanding how different species strategies covary with early-season risks, and how such risks may shift with continued climate change. For temperate zones, many species strategies appear designed to cope with frost, either via avoidance or tolerance \citep{alberto11, lenz16, allevato19}, as utilizing the early portion of the growing season can be especially critical for species in colder regions \citep{morin07, dantec15}. Climate change at once shifts spring phenology---thus potentially changing each species frost risk---but also appears to reshape the drivers of frost climatically \citep{cat2021pep}. Layered onto this are possible shifts in early-season herbivory with warming \citep{meineke2019}, and trade-offs in early-season risks with access to a long growing season for growth and reproduction, with some species avoiding competition through being very early \citep{guo22}. Common garden studies that track and report frost and herbivory damage, alongside timing of flowering and fruiting, could help to understand the complex fitness landscape of spring phenology.
